---
title: "Predicting Bitcoin and Etherium prices using ARIMA and VAR"
author: "Michał Szałanski"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	warning = FALSE,
	comment = '      ',
	fig.width = 16,
	fig.height = 9
)
```

# Abstract
For predicting Bitcoin and Etherium prices both ARIMA and VAR models 
seem to be good choices. ARIMA was slightly weaker with sMAPE of
*4.3%*, compared to *3.4%* achieved by the VAR model. It was observed 
that, for the examined period of time, the Bitcoin and Etherium 
prices were cointegrated. 


# Aim of the study
This study tries to create optimal models for predicting Bitcoin and Etherium
prices by testing currency cointegration and creating optimized ARIMA and VAR models.

# Data

This study will be conducted on the daily closing prices of two cryptocurrencies - 
Bitcoin and Etherium. Bitcoin was chosen because it is a dominant force on the market, 
Etherium was chosen because it seems to follow Bitcoin price very closely.   

The dates are from 2019-01-01 to today. This gives enough data to estimate the models,
while at the same time removes the one-of-a-kind huge price surge that happened in 
2018.   

Out of sample period for estimation was chosen to be 30 days. It should be enough
to test the models, but not too long as to give increased errors.   

The data structure is as follows. It was downloaded from coinmarketcap.com and saved into csv files.   

```{R libraries, message = FALSE, warning = FALSE, error = FALSE}
### Preparation
# Libraries
library(tidyverse)
library(ggplot2) # nice plots
library(ggthemes) 
library(ggtech) # For the airbnb theme

library(xts)
library(lmtest)
library(tseries)
library(urca)
library(fUnitRoots)
library(quantmod)
library(fBasics)
library(forecast)
library(vars)
library(Metrics)

# For better plotting
theme_set(theme_airbnb_fancy())
orange = "#FFB400"
purple = "#7B0051"

# Fix for the month shorthands
invisible(Sys.setlocale("LC_ALL","English"))

### Data downloading
btc <- read_csv('data/btc.csv')
eth <- read_csv('data/eth.csv')

source("functions.R")

btc %>% tail
```

\newpage

Bitcoin prices exhibit significant changes in price in both directions, with no 
visible trend. None of the changes come close to these observed in 2018.   

```{r plot_btc}
btc %>%
	ggplot(aes(Date, Close)) +
	geom_area(fill = orange, color = orange, alpha = 0.5) +
	labs(title = "Bitcoin historical prices") +
	theme(
		text = element_text(family = "sans"), # Remove the font warnings
		plot.title = element_text(family = "sans")
	)
```

Etherium prices looks very similar to Bitcoin, although 
they seem to have slightly higher variance.   

```{r plot_eth}
eth %>%
	ggplot(aes(Date, Close)) +
	geom_area(fill = orange, color = orange, alpha = 0.5) +
	labs(title = "Etherium historical prices") +
	theme(
		text = element_text(family = "sans"), # Remove the font warnings
		plot.title = element_text(family = "sans")
	)
```

\newpage

Plotting both prices as % of a max price confirms this. 
Etherium reacts to changes stronger than Bitcoin. They do however 
seem very cointegrated. 

```{r plot_both}
btc$source <- 'Bitcoin'
eth$source <- 'Etherium'

btc$ClosePercent <- btc$Close / max(btc$Close)
eth$ClosePercent <- eth$Close / max(eth$Close)


bind_rows(btc, eth) %>%
	ggplot(aes(Date, 
			   ClosePercent,
			   color = source)) +
	geom_line() +
	scale_color_manual(values = c(orange, purple)) +
	labs(title = "Close price as % of max price for each coin") +
	theme(
		text = element_text(family = "sans"), # Remove the font warnings
		plot.title = element_text(family = "sans")
	)
```

This is the same plot, but as a difference between prices. 
As it can be seen, when the change is very short and strong (e.g. around 2020-03), 
the prices move like one. This could be a result of automated 
trading, but more in-depth research would be required.   

```{r plot_both_2}
bind_rows(btc, eth) %>%
	ggplot(aes(Date, ClosePercent)) +
	geom_line(color = orange) +
	labs(title = "Close price as % of max price for each coin",
		 subtitle = 'Colored only the difference') +
	theme(
		text = element_text(family = "sans"), # Remove the font warnings
		plot.title = element_text(family = "sans")
	)
```

```{r }
data <- data.frame(
	date = btc$Date,
	btc  = btc$Close,
	eth  = eth$Close
)

remove(btc)
remove(eth)  

data$Dbtc <- diff.xts(data$btc)
data$Deth <- diff.xts(data$eth)

# Drop the first row with NAs
data <- data %>% drop_na()

# Extract last 30 days for predictions
data_test <- data %>% tail(30)

# Remove them to avoid look ahead bias
data <- data %>% head(-30) # So usefull
```

\newpage
# ARIMA - Bitcoin

Breusch–Godfrey test signifies a autocorrelation of order 1, so 
at least one augmentation is needed.

```{r }
testdf(variable = data$btc,
	   max.augmentations = 5)
```

After differencing, all Breusch–Godfrey tests are insignificant. This 
means that the price of Bitcoin is a I(1) process. 

```{r }
testdf(variable = data$Dbtc,
	   max.augmentations = 5)
```

Some lags seem to be on the verge of being significant. Thanks to the
power of modern PC, running *auto.arima* on max parameters of (240, 1, 240) 
takes only a minute. 

```{r plot_arima_btc}
par(mfrow = c(2, 1)) 
acf(data$Dbtc,
	lag.max = 36, 
	ylim = c(-0.1, 0.1),    
	lwd = 5,               
	col = orange) 

pacf(data$Dbtc, 
	 lag.max = 36, 
	 lwd = 5,
	 col = orange)
par(mfrow = c(1, 1))
```

\newpage

This is the final ARIMA(20, 1, 0) model, estimated after auto.arima 
and iterative removal of insignificant variables.  

```{r armia_btc}
BTC.arima2010.small <- Arima(data$btc,  
					  		order = c(20, 1, 0),
							fixed = c(
								NA, #1
								rep(0, 7),
								NA, #9
								rep(0, 8),
								NA, # 18
								0,
								NA # 20
							)
)
coeftest(BTC.arima2010.small)
```

The residuals have a mean of zero, but the variance is not
constant. This is due to the changes in variance of the 
underlying variable.   

```{r fig.height = 6}
plot(resid(BTC.arima2010.small))
```
Nothing is significant, which means that the 
residuals are not autocorrelated.

```{r double_plot_arima, fig.height = 8}
par(mfrow = c(2, 1)) 
acf(resid(BTC.arima2010.small), 
	lag.max = 36,
	ylim = c(-0.1, 0.1), 
	lwd = 5, col = orange,
	na.action = na.pass)
pacf(resid(BTC.arima2010.small), 
	 lag.max = 36, 
	 lwd = 5, col = orange,
	 na.action = na.pass)
par(mfrow = c(1, 1))

```

\newpage

BJ test confirms that the residuals are not autocorrelated. 

```{r }
Box.test(resid(BTC.arima2010.small), type = "Ljung-Box", lag = 10)
```

\newpage

# ARIMA - Etherium

Judging by the Breusch–Godfrey test, Etherium prices are a I(1) process, so one 
augmentation should be enough.

```{r }
testdf(variable = data$eth,
	   max.augmentations = 5)
```

Testing the variable differenced once confirms this. Breusch–Godfrey is not 
significant here for 0 augmentations, which means that there is 
no autocorrelation in the new variable.

```{r }
testdf(variable = data$Deth,
	   max.augmentations = 5)
```

Some lags seem to be significant, so *auto.arima* was run on max parameters of (240, 1, 240).

```{r plot_arima_eth}
par(mfrow = c(2, 1)) 
acf(data$Deth,
	lag.max = 36, 
	ylim = c(-0.1, 0.1),    
	lwd = 5,               
	col = orange) 

pacf(data$Deth, 
	 lag.max = 36, 
	 lwd = 5,
	 col = orange)
par(mfrow = c(1, 1)) 
```

\newpage

This is the final model, taken form auto.arima, with insignificant 
variables iteratively removed. It is ARIMA(5, 1, 20).

```{r arima_eth}
ETH.arima5120.small <- Arima(data$eth,  
							order = c(5, 1, 20),
							fixed = c(
								rep(0,3),
								NA,
								NA,
								rep(0, 3),
								NA,
								NA,
								rep(0, 3),
								NA,
								0,
								0,
								rep(0, 8),
								NA
							)
)

coeftest(ETH.arima5120.small)
```

Residuals have a mean of 0, but they are not stationary. 
This is because of the changes in variance. This is due to the original data that 
has significant spikes in price in some places, so nothing more can be done 
about it.   

```{r fig.height = 6}
plot(resid(ETH.arima5120.small))
```

Everything is insignificant, or on the verge of significance, which is good news.   

```{r double_plot_eth, fig.height = 7}
par(mfrow = c(2, 1)) 
acf(resid(ETH.arima5120.small), 
	lag.max = 36,
	ylim = c(-0.1, 0.1), 
	lwd = 5, col = orange,
	na.action = na.pass)
pacf(resid(ETH.arima5120.small), 
	 lag.max = 36, 
	 lwd = 5, col = orange,
	 na.action = na.pass)
par(mfrow = c(1, 1))

```

Box-Ljung test confirms that the residuals are not in fact
autocorrelated. 

```{r }
Box.test(resid(ETH.arima5120.small), type = "Ljung-Box", lag = 10)
```

\newpage
# ARIMA forecasting

30 day ARIMA forecast for the Bitcoin is close to the real values,
although it seems to overestimate the price and ignore the 
downward trend present in the data. 
Because the price is so stable, the predicted values looks kind of 
like a naive prediction.    

```{r arima_forecast_btc}
forecast(BTC.arima2010.small, h = 30) %>%                  # Forecast
	{data.frame(f_mean  = as.numeric(.$mean),              # Make into a DF
			   f_lower = as.numeric(.$lower[, 2]),
			   f_upper = as.numeric(.$upper[, 2]))} %>%
	cbind(data_test) %>%                                   # Add the real data
	ggplot(aes(x = date)) +                                # Plot
		geom_line(aes(y = btc), color = purple) +
		geom_line(aes(y = f_mean), color = orange) +
		geom_line(aes(y = f_lower), color = 'red') +
		geom_line(aes(y = f_upper), color = 'red') +
		labs(title = "30 day prediction of BTC",
			 subtitle = 'Yellow is the prediction') +
		theme(
			text = element_text(family = "sans"), # Remove the font warnings
			plot.title = element_text(family = "sans")
		)
```

Similar results for Etherium - although here it seems that 
the model was able to predict the general downward trend.
It can be said that it predicted it too much.    

```{r arima_forecast_eth}
forecast(ETH.arima5120.small, h = 30) %>%                  # Forecast
	{data.frame(f_mean  = as.numeric(.$mean),              # Make into a DF
				f_lower = as.numeric(.$lower[, 2]),
				f_upper = as.numeric(.$upper[, 2]))} %>%
	cbind(data_test) %>%                                   # Add the real data
	ggplot(aes(x = date)) +                                # Plot
	geom_line(aes(y = eth), color = purple) +
	geom_line(aes(y = f_mean), color = orange) +
	geom_line(aes(y = f_lower), color = 'red') +
	geom_line(aes(y = f_upper), color = 'red') +
	labs(title = "30 day prediction of ETH",
		 subtitle = 'Yellow is the prediction') +
	theme(
		text = element_text(family = "sans"), # Remove the font warnings
		plot.title = element_text(family = "sans")
	)
```


# Cointegration testing

Since Bitcoin and Etherium are both integrated of 
order I(1), they could be cointegrated with each other.    

To test this cointegration, a simple 
linear model can be estimated. The coefficient for eth 
is very significant, which means that they are in fact cointegrated.    

It seems that a increase in price of ETH by 1 USD means an
increase of BTC price by around 37 USD.    

```{r lm}
# BTC and ETH are I(1)
model.coint <- lm(btc ~ eth, 
				  data = data)

summary(model.coint)
```

But, are the residuals autocorrelated?   

Yes, they are autocorrelated of order one. This means that lagged 
residuals need to be included in the model.  

```{r }
testdf(variable = residuals(model.coint), 
	   max.augmentations = 3)

```

The inclusion of lagged residuals seem to help. The coefficient is significant, which 
hints that there is a long term relationship
between this variables in addition to the short term one
(indicated by eth).   

```{r better_lm}
data$lresid <- lag.xts(residuals(model.coint))
model.ecm <- lm(btc ~ eth + lresid,
				data = data) 
summary(model.ecm)
```

Granger test confirms that, for 5 lags, there exist a 
Granger Causality eth -> btc.   

```{r }
grangertest(btc ~ eth, 
			data = data,
			order = 5) 
```

The same results for the other direction. 
Significant Granger Causality for btc -> eth.   

```{r }
grangertest(eth ~ btc, 
			data = data,
			order = 5)
```


\newpage
# VAR model
*VARselect* function is very useful when searching for the right number of lags in the VAR model.
Here, the optimal number seem to be 7 lags, based on AIC and FPE. Could also be 1, 
but that is not very interesting. 
```{r var}
VARselect(data[ , c('btc', 'eth')], # input data for VAR
		  lag.max = 10)     # maximum lag
```

A large number of variables are significant for p < 0.05.
Especially lags 3 and 4 seem to be important for this model. 


```{r }
var.model <- VAR(data[ , c('btc', 'eth')], p = 7)
summary(var.model)
```

Portmanteau test is on the verge of significance,
so another test is needed.

```{r }
serial.test(var.model)
```

Breusch-Godfrey test finds no reason to 
reject the null hypothesis that the residuals 
are not autocorrelated.

```{r }
serial.test(var.model, type = "BG")
```

\newpage

# Forecasting VAR
The 30 day forecast for Bitcoin using the VAR model is very close to 
the real values. Because the currency did not move significantly 
the predictions are also very stable. In some sense they 
even look like a simple linear regression, but this is only due to the previous prices.     

```{r forecast_var_btc}
# Forecast for 30 days
var.forecast <- predict(var.model, n.ahead = 30, ci = 0.95) 

# Bitcoin
var.forecast[["fcst"]][["btc"]] %>%
	as.data.frame() %>%
	dplyr::select(f_mean = fcst,
		   		  f_lower = lower,
		          f_upper = upper) %>%
	cbind(data_test) %>%                                   # Add the real data
	ggplot(aes(x = date)) +                                # Plot
	geom_line(aes(y = btc), color = purple) +
	geom_line(aes(y = f_mean), color = orange) +
	geom_line(aes(y = f_lower), color = 'red') +
	geom_line(aes(y = f_upper), color = 'red') +
	labs(title = "30 day prediction of BTC",
		 subtitle = 'Yellow is the prediction') +
	theme(
		text = element_text(family = "sans"), # Remove the font warnings
		plot.title = element_text(family = "sans")
	)
```


Same story with Etherium. The predictions are very stable, but they do follow the 
general trend in the data. They also look kind of like a linear regression.      


```{r forecast_var_eth}
var.forecast[["fcst"]][["eth"]] %>%
	as.data.frame() %>%
	dplyr::select(f_mean = fcst,
				  f_lower = lower,
				  f_upper = upper) %>%
	cbind(data_test) %>%                                   # Add the real data
	ggplot(aes(x = date)) +                                # Plot
	geom_line(aes(y = eth), color = purple) +
	geom_line(aes(y = f_mean), color = orange) +
	geom_line(aes(y = f_lower), color = 'red') +
	geom_line(aes(y = f_upper), color = 'red') +
	labs(title = "30 day prediction of ETH",
		 subtitle = 'Yellow is the prediction') +
	theme(
		text = element_text(family = "sans"), # Remove the font warnings
		plot.title = element_text(family = "sans")
	)
```

\newpage

# Forecast comparison

Metric used here is sMAPE, Symmetric Mean Absolute Percentage Error. 
It was chosen because it is unit indifferent and it calculates 
negative errors with the same weight as the positive ones.     

The results are good, sMAPE of *4.3%* for ARIMA and *3.4%* for
VAR are very solid. It can be seen that for this data, 
the VAR model performed significantly better than ARIMA.   

The differences between BTC and ETH within both models 
were insignificant, which hints that these type of models 
could be used with other cryptocurrencies. 

```{r forecast_comparison}
forecast.arima.btc <- forecast(BTC.arima2010.small, h = 30) %>% 
	.$mean %>%
	as.vector()

forecast.arima.eth <- forecast(ETH.arima5120.small, h = 30) %>% 
	.$mean %>%
	as.vector()

forecast.var.btc <- var.forecast[["fcst"]][["btc"]] %>%
	as.data.frame() %>%
	.$fcst

forecast.var.eth <- var.forecast[["fcst"]][["eth"]] %>%
	as.data.frame() %>%
	.$fcst

forecast.comparison <- data.frame(
	btc_real = data_test$btc,
	btc_arima = forecast.arima.btc,
	btc_var = forecast.var.btc,
	eth_real = data_test$eth,
	eth_arima = forecast.arima.eth,
	eth_var = forecast.var.eth
)

data.frame(
	ARIMA_BTC = smape(forecast.comparison$btc_real, forecast.comparison$btc_arima),
	ARIMA_ETH = smape(forecast.comparison$eth_real, forecast.comparison$eth_arima),
	VAR_BTC   = smape(forecast.comparison$btc_real, forecast.comparison$btc_var),
	VAR_ETH   = smape(forecast.comparison$eth_real, forecast.comparison$eth_var)
)
```


# Conclusion

Based on the observed period, it seems that both ARIMA and VAR models 
are good choices for predicting the price of Bitcoin and Etherium. 
Final models for ARIMA were:   

- ARIMA(20, 1, 0) - Bitcoin
- ARIMA(5, 1, 20) - Etherium

For VAR, the best number of lags was 7. The VAR model achieved better results,
thanks to very strong cointegration of the two prices.   

Next steps would be to test more currency pairs (ideally using some 
kind of automation automation). If good results were found 
for some smaller currencies, they could be used for trading. 
When testing multiple times, one needs to be wary of the type one error.
Assuming a lower p value of e.g. 0.01 would be advisable.   

Other possibilities could be to use cross validation to test the performance 
of the models on different parts of data (the one chosen was quite calm). 

